Structuring Theories with Implicit Morphisms

### 2018-10-13 implicit.tex
remove module expressions and unions

\begin{example}\label{ex:dg2gimplicit}
Going back to the morphism presented in \autoref{ex:dg2g}, we simply add the keyword \lstinline|implicit| in front, to declare it to be implicit:
\begin{mmtcode}
implicit view DG2G : ?DivisionGroup -> ?Group =
  universe = universe ❙
  div = [a,b] a ∘ (b⁻¹) ❙
  nonEmpty = e ❙
  // axioms ...❙
❚
\end{mmtcode}
\end{example}

@@

\paragraph{Obtaining Implicit Morphisms}
Based on the above attributes, we define the set $\IMo{S}{T}\sq\Mo{S}{T}$ of \textbf{implicit} morphisms to contain the following elements:
\begin{itemize}
 \item all declared morphisms $m:S\to T=\{\sigma\}$ whose declaration carries the attribute $\keyword{implicit}$,
 \item all morphisms that additional modularity principles designate as implicit based on the use of additional attributes.
\end{itemize}

To define well-formedness for our extended syntax, we say that adding an attribute to a declaration is only well-formed if there is (up to equality of morphisms) at most one implicit morphism for any pair of theories, i.e., $\IMo{S}{T}$ has cardinality $0$ or $1$.
Depending on what modularity principles we add, equality of morphisms may or may not be decidable.
Therefore, in the general case, implementations may have to employ sound but incomplete criteria for equality (see Sect.~\ref{sec:commute}).

The following example phrases the restriction on implicit morphisms in categorical terms:

\begin{example}[Category of Theories and Implicit Morphisms (continued from Ex.~\ref{sem:cat})]\label{impl:cat}
We define the following morphisms are implicit:
\begin{compactitem}
 \item all identity morphisms $\id{T}$,
 \item all compositions $M;N$ of implicit morphisms $M$ and $N$.
\end{compactitem}

Thus, the implicit morphisms form a thin broad subcategory of theories and morphisms.
In other words, the implicit morphisms form commutative diagram.
\end{example}

@@

\begin{example}
	For our morphism $\cn{DG2G}$ from \autoref{ex:dg2gimplicit}, this means that all symbols declared in the theory $\cn{DivisionGroup}$ (see \autoref{ex:thgroup}) are now visible in the theory $\cn{Group}$ with the definitions provided by $\cn{DG2G}$.
\end{example}

It is plausible to say that there are implicit morphisms $S_i\harr S_1\cup S_2$ for $i=1,2$ that map all constants to themselves.
However, union expressions do not interact very well with implicit morphisms.
First of all, we have to be careful to maintain the uniqueness condition.
For example, assume there is a implicit morphism $M:S\to T$ that is not simply an inclusion.
Then the union $S\cup T$ cannot be well-formed.
If it were, we would have two different implicit morphisms $S\to S\cup T$, namely $M$ and the inclusion into the union.

@@

Now to look up the implicit morphism (if any) from $s_1\cup\ldots\cup s_k\to t_1\cup\ldots\cup t_l$, we look up for every $i$ an implicit morphism $m_i:s_i\to t_j$ for some $j$.
If all these $m_i$ exist and $m=m_1\cup \ldots\cup m_k$ is well-formed, then $m$ is the needed implicit morphism.
\end{example}

@@

\ednote{@DM: add the isomorphism of DG2G}
\begin{example}\label{group:iso}
We give the morphism $\cn{G2DG}$.
\end{example}

>>>>

\begin{example}\label{ex:dg2gimplicit}
We can now change the declaration of the morphism $\cn{DG2G}$ from Ex.~\ref{ex:dg2g} by adding the attribute \keyword{implicit}.
\end{example}

@@

\paragraph{Obtaining Implicit Morphisms}
Based on the above attributes, we define the set $\IMo{S}{T}\sq\Mo{S}{T}$ of \textbf{implicit} morphisms to contain the following elements:
\begin{itemize}
 \item all declared morphisms $m:S\to T=\{\sigma\}$ whose declaration carries the attribute $\keyword{implicit}$,
\begin{nomodexp}
 \item all identity morphisms $\id{T}$,
 \item all compositions $M;N$ of implicit morphisms $M$ and $N$,
\end{nomodexp}
 \item all morphisms that additional language features designate as implicit based on the use of additional attributes.
\end{itemize}

To define well-formedness for our extended syntax, we say that adding an attribute to a declaration is only well-formed if there is (up to equality of morphisms) at most one implicit morphism for any pair of theories, i.e., $\IMo{S}{T}$ has cardinality $0$ or $1$.
Thus, the implicit morphisms form a thin broad subcategory of theories and morphisms.
In other words, the implicit morphisms form commutative diagram.

\begin{modexp}
Depending on what modularity principles we add, equality of morphisms may or may not be decidable.
Therefore, in the general case, implementations may have to employ sound but incomplete criteria for equality (see Sect.~\ref{sec:commute}).
\end{modexp}

\begin{modexp}
The following example phrases the restriction on implicit morphisms in categorical terms:
\begin{example}[Category of Theories and Implicit Morphisms (continued from Ex.~\ref{sem:cat})]\label{impl:cat}
We define the following morphisms are implicit:
\begin{compactitem}
 \item all identity morphisms $\id{T}$,
 \item all compositions $M;N$ of implicit morphisms $M$ and $N$.
\end{compactitem}
Thus, the implicit morphisms form a thin broad subcategory of theories and morphisms.
In other words, the implicit morphisms form commutative diagram.
\end{example}
\end{modexp}

@@

\begin{example}
	For our morphism $\cn{DG2G}$ from Ex.~\ref{ex:dg2gimplicit}, this means that all symbols declared in the theory $\cn{DivisionGroup}$ (see Ex.~\ref{ex:thgroup}) are now visible in the theory $\cn{Group}$ with the definitions provided by $\cn{DG2G}$.
\end{example}

\begin{union}
It is plausible to say that there are implicit morphisms $S_i\harr S_1\cup S_2$ for $i=1,2$ that map all constants to themselves.
However, union expressions do not interact very well with implicit morphisms.
First of all, we have to be careful to maintain the uniqueness condition.
For example, assume there is a implicit morphism $M:S\to T$ that is not simply an inclusion.
Then the union $S\cup T$ cannot be well-formed.
If it were, we would have two different implicit morphisms $S\to S\cup T$, namely $M$ and the inclusion into the union.

@@

Now to look up the implicit morphism (if any) from $s_1\cup\ldots\cup s_k\to t_1\cup\ldots\cup t_l$, we look up for every $i$ an implicit morphism $m_i:s_i\to t_j$ for some $j$.
If all these $m_i$ exist and $m=m_1\cup \ldots\cup m_k$ is well-formed, then $m$ is the needed implicit morphism.
\end{example}
\end{union}

@@

\ednote{@DM: add the isomorphism of DG2G}
\begin{example}\label{group:iso}
We give the morphism $\cn{G2DG}$.

\end{example}

### 2018-10-13 mmt.tex
remove module expressions and unions

\paragraph{Modularity in MMT}
As much as possible, theories and morphisms are treated uniformly, and we use the word \textbf{module} to refer to them collectively.
\mmt allows constructing large modules from small ones in two ways:
\begin{compactitem}
  \item \textbf{structured modules} contain declarations that import other, previously defined ones,
  \item \textbf{module expressions} are anonymous expressions (akin to formulas, terms, etc.) that denote modules.
\end{compactitem}

These two kinds of constructions have very different advantages but the same expressivity.
For example, we can construct the theory of commutative groups as the structured theory $\cn{CommutativeGroup}=\{\icl{Group},\icl{CommutativeMagma}\}$ or as the theory expression $\cn{Group}\cup\cn{CommutativeMagma}$.
The former is more practical when building named toplevel theories in software systems because each include is a separate declaration that can be parsed, checked, and flattened individually.
While negligible in this example, this becomes relevant quickly in large theories with complex structure.
The latter is more practical when building anonymous theory that are only needed temporarily.
In software systems, they also allow reusing operations like equality and $\lambda$-abstraction that already be implemented at the level of the base logic.
Building large morphism is as important as building large theories: for example, we want to be able to build morphisms out of $\cn{CommutativeGroup}$ by combining compatible morphisms out of $\cn{Group}$ and $\cn{CommutativeMagma}$.

@@

An \mmt \textbf{diagram} consists of a set of named structured module declarations and induces sets of anonymous module expressions.
These are mutually recursive: the names of the former provide the base cases for inductively forming the latter, and the latter may occur in the bodies of the former.
Every diagram induces sets
\begin{compactitem}
 \item $\thy$ of identifiers of theories $t$,
 \item $\Thy\supseteq \thy$ of theory expressions $T$,
 \item $\mo{S}{T}$ of identifiers of atomic morphisms $m$ from $S$ to $T$, and
 \item $\Mo{S}{T}\supseteq\mo{S}{T}$ of morphism expressions $M$ from $S$ to $T$.
\end{compactitem}
Then flattening assigns
\begin{compactitem}
 \item to each $T\in\Thy$ the flat theory $\flt{T}$,
 \item to each $M\in\Mo{S}{T}$ the morphism $\flt{M}$ from $\flt{S}$ to $\flt{T}$.
\end{compactitem}

@@

\begin{definition}[Theory]\label{def:theory}
A \textbf{theory} $T$ is of the form
\begin{grammar}
TDec     & t=\{Dec,\ldots,Dec\}  & \text{theory declaration} \\
Dec      & n:o\opt{=o}           & \text{constant declaration}\\
c        & t?n                   & \text{qualified constant identifiers} \\
o        & c \alt \ldots         & \text{expressions built from constants} \\
T        & t                     & \text{theory expressions}
\end{grammar}

@@

\begin{example}\label{ex:thgroup}
The (flat) theories $\cn{Group}$ and $\cn{DivGroup}$ from the introduction are:
\begin{mmtcode}
Group =
  universe	: type	 # U
  op		: U ⟶ U ⟶ U	 # 1 ∘ 2 
  unit		: U
  inverse : U ⟶ U   # 1 ⁻¹

  axiom_assoc		: ⊦ ∀[x]∀[y]∀[z] x ∘ (y ∘ z) ≐ (x ∘ y) ∘ z 
  axiom_right_unit	: ⊦ ∀[x] x ∘ unit ≐ x 
  axiom_left_unit	: ⊦ ∀[x] unit ∘ x ≐ x 
  axiom_inverse : ⊦ ∀[x] x ∘ (x⁻¹) ≐ unit
\end{mmtcode}
\begin{mmtcode}
DivisionGroup =
  universe : type  # U prec 1 
  div : U ⟶ U ⟶ U  # 1 / 2 
  unit : U  # e prec 1 
  // axioms ... 
\end{mmtcode}
Here we use \lstinline|# <not>| to indicate notations; these are part of the language supported by \mmt but omitted from the formal grammar above because we only need them here to present legible examples.

@@

\begin{definition}[Morphism]\label{def:morphism}
A \textbf{morphism} is of the form
\begin{grammar}
MDec   & m:T\to T=\{Ass,\ldots,Ass\}     & \text{flat morphism declaration}\\
Ass    & c:=o                            & \text{assignment to symbol}\\
M      & m                               & \text{morphism expressions}
\end{grammar}
such that a morphism declaration contains exactly one assignment $c:=o$ for each $c\in\dom{S}$, all of which satisfying $o\in\Exp{T}$.
\end{definition}

\begin{example}\label{ex:dg2g}
	We can now specify the morphisms between our two definitions of groups. The theory of groups as monoids with inverses is given in \autoref{syn:incl}, the theory $\cn{DivGroup}$ and the morphism $\cn{DG2G}$ are as follows:
\begin{mmtcode}
DG2G : DivisionGroup -> Group =
  universe = universe 
  div = [a,b] a ∘ (b⁻¹) 
  unit = unit
  // axioms ...
\end{mmtcode}
The morphism $\cn{DG2G}$ hence maps the universe of a division group to the universe of a group, division to the expression $\lambda a,b.\; a\circ b^{-1}$ and the witness $\cn{nonEmpty}$ of a division group to the unit $e$ of a group. Additionally, the view has to map the axioms of a division group to their \emph{proofs} in $\cn{Group}$ of their translated statements.

The reverse morphism $\cn{G2DG}$ is completely analogous, mapping $\cn{op}$, $\cn{unit}$ and $\cn{inverse}$ to their corresponding definitions in terms of the division operation $\cn{div}$.
\end{example}

@@

\begin{example}[Includes]\label{syn:incl}
We extend the grammar with
\begin{grammar}
  Dec & \icl{S} & \text{include a theory into a theory} \\
  Ass & \icl{M} & \text{include a morphism into a morphism}
\end{grammar}
This allows writing the theory $\cn{Group}$ as in \autoref{ex:thgroup} by extending a theory of monoids:
\begin{mmtcode}
Monoid =
  universe	: type		 # U 
  op		: U ⟶ U ⟶ U	 # 1 ∘ 2 
  unit		: U
  axiom_assoc		: ⊦ ∀[x]∀[y]∀[z] x ∘ (y ∘ z) ≐ (x ∘ y) ∘ z 
  axiom_right_unit	: ⊦ ∀[x] x ∘ unit ≐ x 
  axiom_left_unit	: ⊦ ∀[x] unit ∘ x ≐ x

Group =
  include Monoid
  inverse : U ⟶ U   # 1 ⁻¹ 
  axiom_inverse : ⊦ ∀[x] x ∘ (x⁻¹) ≐ unit
\end{mmtcode}
\end{example}

\begin{example}[Union]\label{syn:union}
We extend the grammar with
\begin{grammar}
  T   & T\cup T & \text{union of theories} \\
  M   & M\cup M & \text{union of morphisms}
\end{grammar}
\end{example}

\begin{example}[Category of Theories]\label{syn:cat}
To obtain the category structure of theories and morphisms, we need identity and composition.
We add them to the grammar as follows:
\begin{grammar}
  M   & \id{T} & \text{identity morphism} \\
  M   & M;M    & \text{morphism composition}
\end{grammar}
\end{example}

@@

\paragraph{Flattening}
Just like the syntax, the definition of flattening is \textbf{compositional} in the sense that new modularity principles can be added later independently of each other.
In particular, for every module expression $X$, we define its flattening $\flt{X}$ by induction on the syntax.
This follows the principle of compositional semantics, e.g., $\flt{(f(X,Y))}$ depends only on $\flt{X}$ and $flt{Y}$.

@@

At this point, we only have one case for declarations $D$, namely constant declarations.
Their flattening is trivial:
 \[\fltd{\Sigma}{n:E[=e]}=\{t?n:E[=e]\}\]
where $t$ is the name of the containing theory.

Correspondingly, for a declared morphism $m:S\to T=\{\sigma\}$, we have $m\in\Mo{S}{T}$ and $\flt{m}$ is defined by induction on the assignments in $\sigma$:
 \[\flt{\cdot}=\es\]
 \[\flt{(\sigma,\;A)}=\flt{\sigma}\cup\fltd{\flt{\sigma}}{A}\]
with the trivial base case
 \[\fltd{\sigma}{c:=e}=\{c:=e\}\]

\begin{example}[Includes (continued from Ex.~\ref{syn:incl})]\label{sem:incl}
Includes add new declarations $D$, so we have to add cases to the definition of $\fltd{\Sigma}{D}$.
We do this as follows
  \[\fltd{\Sigma}{\icl{S}} = \flt{S}\]
This has the effect of copying over all declarations from the included into the including theory.

@@

Also note, that because $\flt{S}$ is a \textit{set} of declarations, the include relation is transitive: if $t$ includes $s$ via two different paths, $\flt{t}$ only contains one copy of the declarations of $\flt{s}$.
The situation is slightly more complicated for morphisms: if a morphism out of $t$ includes two different morphisms out of $s$, these have to agree.
Therefore, flattening and well-formedness must be distinguished:
  \[\fltd{\sigma}{\icl{M}} = \flt{M}\]
under the condition $\flt{M}$ agrees with $\flt{\sigma}$ on any constant that is in the domain of both.

@@

\begin{example}[Union (continued from Ex.~\ref{syn:union})]\label{sem:union}
We define the semantics of unions of theories as
\[\flt{(S\cup T)}=\flt{S}\cup\flt{T}\]
The handling of name clashes is the same as for includes:
Because the theories declared in a diagram have unique names and declarations in the flattening are qualified by theory names, union theories cannot have name clashes.
But we have to watch out when taking unions of morphisms:
We define $M_1\cup M_2\in \Mo{S_1\cup S_2}{T}$ if $M_i\in\Mo{S_i}{T}$ and $M_1$ and $M_2$ agree for every constant in $\dom{\flt{S_1}}\cap\dom{\flt{S_2}}$.
In that case, we put
\[\flt{(M\cup N)}=\flt{M}\cup\flt{N}\]
\end{example}

Finally, we define the semantics of the category structure:
\begin{example}[Category of Theories (continued from Ex.~\ref{syn:cat})]\label{sem:cat}
For $T\in\Thy$, we put $\id{T}\in\Mo{T}{T}$ and define
\[\flt{\id{T}}=\{c:=c \;|\; c:E\in\dom{\flt{T}}\}\]

For $R,S,T\in\Thy$ and $M\in\Mo{R}{S}$ and $N\in\Mo{S}{T}$, we put $M;N\in\Mo{R}{T}$ and define
\[\flt{M;N}=\{c:=\flt{N}(\flt{M}(c)) \;|\; c:E\in\dom{\flt{R}}\}\]

Note that in both cases, we only have to consider constants $c:E$ without definiens.
\end{example}

@@

\subsection{Semantic Properties}

@@

\begin{example}[Union (continued from Ex.~\ref{sem:union})]\label{rel:union}
We have the following soundness theorem for unions: $S_i\harr S_1\cup S_2$.
Moreover, $S_1\cup S_2$ is a colimit in the category of theories and morphism, and $M_1\cup M_2\in\Mo{S_1\cup S_2}{T}$ is the universal morphism out of it.
\end{example}

\begin{example}[Category of Theories (continued from Ex.~\ref{sem:cat})]\label{rel:cat}
Identity and composition are congruent with respect to $\equiv$.
Moreover, composition is associative and identity neutral with respect to $\equiv$.
Thus, each diagram indeed yields the structure of a category with objects $\Thy$ and arrows $\Mo{S}{T}$.
\end{example}

>>>>

\begin{modexp}
\paragraph{Modularity in MMT}
As much as possible, theories and morphisms are treated uniformly, and we use the word \textbf{module} to refer to them collectively.
\mmt allows constructing large modules from small ones in two ways:
\begin{compactitem}
  \item \textbf{structured modules} contain declarations that import other, previously defined ones,
  \item \textbf{module expressions} are anonymous expressions (akin to formulas, terms, etc.) that denote modules.
\end{compactitem}

These two kinds of constructions have very different advantages but the same expressivity.
For example, we can construct the theory of commutative groups as the structured theory $\cn{CommutativeGroup}=\{\icl{Group},\icl{CommutativeMagma}\}$ or as the theory expression $\cn{Group}\cup\cn{CommutativeMagma}$.
The former is more practical when building named toplevel theories in software systems because each include is a separate declaration that can be parsed, checked, and flattened individually.
While negligible in this example, this becomes relevant quickly in large theories with complex structure.
The latter is more practical when building anonymous theory that are only needed temporarily.
In software systems, they also allow reusing operations like equality and $\lambda$-abstraction that already be implemented at the level of the base logic.
Building large morphism is as important as building large theories: for example, we want to be able to build morphisms out of $\cn{CommutativeGroup}$ by combining compatible morphisms out of $\cn{Group}$ and $\cn{CommutativeMagma}$.
\end{modexp}

@@

An \mmt \textbf{diagram} consists of a set of named structured module declarations.
\begin{nomodexp}
For a given diagram, we write $\Thy$ for the set of theory names.
And we write $\Mo{S}{T}$ for the set of morphisms defined by
\begin{compactitem}
 \item for every declaration $m:S\to T=\{\sigma\}$, we have $m\in\Mo{S}{T}$,
 \item for every $T\in\Thy$, we have $\id{T}\in\Mo{T}{T}$,
 \item for every $M\in\Mo{R}{S}$ and $N\in\Mo{S}{T}$, we have $M;N\in\Mo{R}{T}$.
\end{compactitem}
$\Thy$ and $\Mo{S}{T}$ form the category of theories and morphisms.
\end{nomodexp}
\begin{modexp}
Every diagram induces sets of anonymous module expressions.
These are mutually recursive: the names of the former provide the base cases for inductively forming the latter, and the latter may occur in the bodies of the former.
Every diagram induces sets
\begin{compactitem}
 \item $\thy$ of identifiers of theories $t$,
 \item $\Thy\supseteq \thy$ of theory expressions $T$,
 \item $\mo{S}{T}$ of identifiers of atomic morphisms $m$ from $S$ to $T$, and
 \item $\Mo{S}{T}\supseteq\mo{S}{T}$ of morphism expressions $M$ from $S$ to $T$.
\end{compactitem}
\end{modexp}
Then flattening assigns
\begin{compactitem}
 \item to each $T\in\Thy$ the flat theory $\flt{T}$,
 \item to each $M\in\Mo{S}{T}$ the flat morphism $\flt{M}$ from $\flt{S}$ to $\flt{T}$.
\end{compactitem}

@@

\begin{definition}[Theory]\label{def:theory}
The grammar for theories and expressions is of the form
\begin{grammar}
TDec     & T=\{Dec,\ldots,Dec\}  & \text{theory declaration} \\
Dec      & n:E\opt{=E}           & \text{constant declaration}\\
c        & t?n                   & \text{qualified constant identifiers} \\
E        & c \alt \ldots         & \text{expressions built from constants} \\
\end{grammar}

@@

\begin{example}\label{ex:thgroup}
The (flat) theories $\cn{Group}$ and $\cn{DivGroup}$ from the introduction are:
\begin{mmtcode}
Group =
  U	: type
  op		: U ⟶ U ⟶ U	 # 1 ∘ 2 
  unit		: U
  inverse : U ⟶ U   # 1 ⁻¹ 
  // axioms omitted
\end{mmtcode}
\begin{mmtcode}
DivisionGroup =
  U : type
  div : U ⟶ U ⟶ U  # 1 / 2 
  unit : U
  // axioms omitted
\end{mmtcode}
Here we use \lstinline|# <not>| to indicate notations; these are part of the language supported by \mmt but omitted from the formal grammar above because we only need them here to present legible examples.

@@

\begin{definition}[Morphism]\label{def:morphism}
A \textbf{morphism} is of the form
\begin{grammar}
MDec   & m:T\to T=\{Ass,\ldots,Ass\}     & \text{flat morphism declaration}\\
Ass    & c:=o                            & \text{assignment to symbol}\\
M      & m \alt \id{T} \alt M;M          & \text{morphism expressions}
\end{grammar}
such that a morphism declaration contains exactly one assignment $c:=o$ for each $c\in\dom{S}$, all of which satisfying $o\in\Exp{T}$.
\end{definition}

\begin{example}[Morphisms]\label{ex:dg2g}
We give the morphism \cn{DG2G} between the theories from Ex.~\ref{syn:incl}:
\begin{mmtcode}
DG2G : DivisionGroup -> Group =
  U := U
  div := [a,b] a ∘ (b⁻¹) 
  unit := unit
  // assignments to axioms omitted
\end{mmtcode}
It maps the universe and unit of a division group to the corresponding notions of a group.
And we have $\cn{DG2G}(a/b)=a\circ b^{-1}$.
Additionally, the morphism must maps every axiom of \cn{DivGroup} to a proof in \cn{Group} of the translated statement, but we omit those assignments.
\end{example}

@@

\begin{example}[Includes]\label{syn:incl}
We extend the grammar with
\begin{grammar}
  Dec & \icl{S} & \text{include a theory into a theory} \\
  Ass & \icl{M} & \text{include a morphism into a morphism}
\end{grammar}
This allows writing the theory $\cn{Group}$ from Ex.~\ref{ex:thgroup} by extending a theory of monoids.
Again omitting all axioms, this looks as follows:
\begin{mmtcode}
Monoid =
  U: type
  op : U ⟶ U ⟶ U	 # 1 ∘ 2 
  unit	: U
Group =
  include Monoid
  inverse : U ⟶ U   # 1 ⁻¹ 
\end{mmtcode}
\end{example}

\begin{union}
\begin{example}[Union]\label{syn:union}
We extend the grammar with
\begin{grammar}
  T   & T\cup T & \text{union of theories} \\
  M   & M\cup M & \text{union of morphisms}
\end{grammar}
\end{example}
\end{union}

\begin{modexp}
\begin{example}[Category of Theories]\label{syn:cat}
To obtain the category structure of theories and morphisms, we need identity and composition.
We add them to the grammar as follows:
\begin{grammar}
  M   & \id{T} & \text{identity morphism} \\
  M   & M;M    & \text{morphism composition}
\end{grammar}
\end{example}
\end{modexp}

@@

\paragraph{Flattening}
Just like the syntax, the definition of flattening is \textbf{compositional} in the sense that new modularity principles can be added later independently of each other.
In particular, for every module expression $X$, we define its flattening $\flt{X}$ by induction on the syntax.
This follows the principle of compositional semantics, e.g., $\flt{f(X,Y)}$ depends only on $\flt{X}$ and $\flt{Y}$.

@@

At this point, we only have one case for declarations $D$, namely constant declarations.
Their flattening is trivial:
 \[\fltd{\Sigma}{(n:E[=e])}=\{t?n:E[=e]\}\]
where $t$ is the name of the containing theory.

Correspondingly, for a declared morphism $m:S\to T=\{\sigma\}$, we have $m\in\Mo{S}{T}$ and $\flt{m}$ is defined by induction on the assignments in $\sigma$:
 \[\flt{\cdot}=\es\]
 \[\flt{(\sigma,\;A)}=\flt{\sigma}\cup\fltd{\flt{\sigma}}{A}\]
with the trivial base case
 \[\fltd{\sigma}{(c:=e)}=\{c:=e\}\]
\begin{nomodexp}
Moreover, for $T\in\Thy$ we define
\[\flt{\id{T}}=\{c:=c \;|\; c:E\in\flt{T}\}\]
And for $M;N\in\Mo{R}{T}$, we define
\[\flt{(M;N)}=\{c:=\flt{N}(\flt{M}(c)) \;|\; c:E\in\flt{R}\}\]
Note that in both cases, we only have to consider $R$-constants without definiens.
\end{nomodexp}

\begin{example}[Includes (continued from Ex.~\ref{syn:incl})]\label{sem:incl}
Includes add new declarations $D$, so we have to add cases to the definition of $\fltd{\Sigma}{D}$.
We do this as follows
  \[\fltd{\Sigma}{(\icl{S})} = \flt{S}\]
This has the effect of copying over all declarations from the included into the including theory.

@@

Also note, that because $\flt{S}$ is a \textit{set} of declarations, the include relation is transitive: if $t$ includes $s$ via two different paths, $\flt{t}$ only contains one copy of the declarations of $\flt{s}$.
The situation is slightly more complicated for morphisms: if a morphism out of $t$ includes two different morphisms out of $s$, these have to agree.
Therefore, flattening and well-formedness must be distinguished:
  \[\fltd{\sigma}{(\icl{M})} = \flt{M}\]
under the condition $\flt{M}$ agrees with $\flt{\sigma}$ on any constant that is in the domain of both.

@@

\begin{union}
\begin{example}[Union (continued from Ex.~\ref{syn:union})]\label{sem:union}
We define the semantics of unions of theories as
\[\flt{(S\cup T)}=\flt{S}\cup\flt{T}\]
The handling of name clashes is the same as for includes:
Because the theories declared in a diagram have unique names and declarations in the flattening are qualified by theory names, union theories cannot have name clashes.
But we have to watch out when taking unions of morphisms:
We define $M_1\cup M_2\in \Mo{S_1\cup S_2}{T}$ if $M_i\in\Mo{S_i}{T}$ and $M_1$ and $M_2$ agree for every constant in $\dom{\flt{S_1}}\cap\dom{\flt{S_2}}$.
In that case, we put
\[\flt{(M\cup N)}=\flt{M}\cup\flt{N}\]
\end{example}
\end{union}

@@

\begin{modexp}
\subsection{Semantic Properties}

@@

\begin{union}
\begin{example}[Union (continued from Ex.~\ref{sem:union})]\label{rel:union}
We have the following soundness theorem for unions: $S_i\harr S_1\cup S_2$.
Moreover, $S_1\cup S_2$ is a colimit in the category of theories and morphism, and $M_1\cup M_2\in\Mo{S_1\cup S_2}{T}$ is the universal morphism out of it.
\end{example}
\end{union}

\begin{example}[Category of Theories (continued from Ex.~\ref{sem:cat})]\label{rel:cat}
Identity and composition are congruent with respect to $\equiv$.
Moreover, composition is associative and identity neutral with respect to $\equiv$.
Thus, each diagram indeed yields the structure of a category with objects $\Thy$ and arrows $\Mo{S}{T}$.
\end{example}
\end{modexp}
